# Survey Weighting: Making Samples Representative

The press release from the polling firm hits your inbox three weeks before Election Day: "Exclusive Poll: Mayor Johnson leads 54-43!" Your editor wants it for the front page.

But you dig into the methodology section buried at the bottom: "Survey of 823 likely voters." You call the polling firm for details. How did they define "likely voters"? How did they reach them?

The pollster explains: "We called landlines during business hours." You pause. Who has landlines anymore? Who's home on a Tuesday afternoon? Retirees, mostly. Young people have cell phones. Working people are at their jobs.

Your sample is skewed. It over-represents older people and under-represents younger ones. If those groups vote differently—and they almost always do—your poll doesn't represent the electorate. It represents the people you happened to reach.

This is where **survey weighting** comes in. It's a mathematical fix for an inevitable problem: your sample never perfectly matches the population you're trying to understand.

## The Fundamental Problem

Real-world surveys always have the same challenge: some types of people respond more than others.

**Who over-responds:**
- Older people (they answer phones)
- Women (higher response rates)
- College graduates (they like surveys)
- White people (demographic reality of sampling frames)
- Politically engaged people (they care about polls)

**Who under-responds:**
- Young people (ignore calls, use cell phones exclusively)
- Minorities (harder to reach, lower trust in institutions)
- Low-income residents (multiple jobs, no time)
- Renters who move frequently (outdated contact info)

If you just count responses, you're hearing disproportionately from some groups and not enough from others. And if those groups have different opinions—they usually do—your results are biased.

## What Weighting Does

Survey weighting adjusts the "volume" on different voices to match the population.

**Simple example:**
- Your sample: 70% women, 30% men
- Actual population: 50% women, 50% men
- Solution: Give each woman's response less weight (count as ~0.71 people), give each man's response more weight (count as ~1.67 people)

Now your weighted sample reflects the actual gender split, even though your raw sample didn't.

## How Weights Work

Each survey respondent gets a **weight** that shows how many people they represent:

- **Weight = 1.0**: This person represents exactly themselves (properly represented in sample)
- **Weight = 0.5**: This person represents half a person (their group is over-represented)
- **Weight = 2.0**: This person represents two people (their group is under-represented)

When analyzing the data, you sum weights instead of counting people.

## Setting Up

```{r setup, message=FALSE}
library(tidyverse)
library(knitr)
```

## Working with Real Survey Data

Let's use data from a 2020 survey about voting behavior. The survey asked people if they planned to vote, and pollsters weighted responses to match population demographics:

```{r load-data, message=FALSE}
nonvoters_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/nonvoters_data.csv")

glimpse(nonvoters_data)
```

**Key variables:**
- `weight`: The survey weight for each respondent
- `Q21`: Voting intention (1=Yes, 2=No, 3=Unsure)
- `voter_category`: Past voting history
- Demographics: race, gender, education, income

## Examining the Weights

Let's look at how the weights are distributed:

```{r examine-weights, fig.width=10, fig.height=6}
# Summary statistics
summary(nonvoters_data$weight)

# Histogram of weights
ggplot(nonvoters_data, aes(x = weight)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 1.0, color = "red", linetype = "dashed", linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Distribution of Survey Weights",
    subtitle = "Red line shows weight = 1.0 (properly represented); most cluster near this",
    x = "Weight",
    y = "Number of Respondents"
  )
```

**What this tells us:**
- Most weights cluster around 1.0 (sample reasonably represents these people)
- Some weights are as low as 0.2 (very over-represented)
- Some weights are as high as 3.0 (very under-represented)

## Interpreting Specific Weights

Here's what different weights mean:

```{r weight-examples}
weight_examples <- tibble(
  Weight = c(0.25, 0.50, 1.00, 1.50, 2.50),
  Interpretation = c(
    "Counts as 1/4 person (group very overrepresented in sample)",
    "Counts as 1/2 person (group moderately overrepresented)",
    "Counts as exactly 1 person (group properly represented)",
    "Counts as 1.5 people (group moderately underrepresented)",
    "Counts as 2.5 people (group very underrepresented)"
  )
)

kable(weight_examples, caption = "What different weights mean")
```

## The Critical Comparison: Weighted vs. Unweighted

Let's analyze the same question two ways and see how much weighting matters.

**Question:** "Do you plan to vote in the upcoming election?"

### Method 1: Unweighted (Just Count People)

This is what you'd get if you ignored the weights—just tally up responses:

```{r unweighted-analysis}
# Overall voting intentions, unweighted
unweighted_overall <- nonvoters_data |>
  filter(!is.na(Q21), Q21 > 0) |>
  count(Q21) |>
  mutate(
    percentage = n / sum(n) * 100,
    response = case_when(
      Q21 == 1 ~ "Yes - Plans to Vote",
      Q21 == 2 ~ "No - Doesn't Plan to Vote",
      Q21 == 3 ~ "Unsure/Undecided"
    )
  ) |>
  select(response, n, percentage)

kable(unweighted_overall, digits = 1,
      caption = "UNWEIGHTED: Simply counting respondents")
```

### Method 2: Weighted (Sum the Weights)

Now let's do it right—sum the weights instead of counting people:

```{r weighted-analysis}
# Overall voting intentions, weighted
weighted_overall <- nonvoters_data |>
  filter(!is.na(Q21), Q21 > 0) |>
  group_by(Q21) |>
  summarize(weighted_n = sum(weight)) |>
  mutate(
    percentage = weighted_n / sum(weighted_n) * 100,
    response = case_when(
      Q21 == 1 ~ "Yes - Plans to Vote",
      Q21 == 2 ~ "No - Doesn't Plan to Vote",
      Q21 == 3 ~ "Unsure/Undecided"
    )
  ) |>
  select(response, weighted_n, percentage)

kable(weighted_overall, digits = 1,
      caption = "WEIGHTED: Summing weights to reflect actual population")
```

### The Difference

Let's compare them side by side:

```{r compare-overall}
comparison <- unweighted_overall |>
  left_join(weighted_overall, by = "response", suffix = c("_unweighted", "_weighted")) |>
  mutate(difference = percentage_weighted - percentage_unweighted) |>
  select(response, percentage_unweighted, percentage_weighted, difference)

kable(comparison, digits = 1,
      col.names = c("Response", "Unweighted %", "Weighted %", "Difference"),
      caption = "Impact of weighting on poll results")
```

Notice the shifts! They might seem small—2-3 percentage points—but in close elections, that's the difference between calling a race and saying it's too close to call.

## Breaking Down by Voter Type

Let's see how weighting affects different categories of voters. Some groups might be more skewed than others.

### Unweighted Results by Category

```{r unweighted-by-category}
unweighted_by_voter <- nonvoters_data |>
  filter(!is.na(Q21), Q21 > 0) |>
  group_by(voter_category, Q21) |>
  summarize(count = n(), .groups = "drop_last") |>
  mutate(
    total = sum(count),
    percentage = count / total * 100
  ) |>
  ungroup()

unweighted_summary <- unweighted_by_voter |>
  pivot_wider(
    id_cols = voter_category,
    names_from = Q21,
    values_from = percentage,
    names_prefix = "Q21_"
  ) |>
  rename(
    "Yes (%)" = Q21_1,
    "No (%)" = Q21_2,
    "Unsure (%)" = Q21_3
  )

kable(unweighted_summary, digits = 1,
      caption = "UNWEIGHTED: Voting intentions by past voting behavior")
```

### Weighted Results by Category

```{r weighted-by-category}
weighted_by_voter <- nonvoters_data |>
  filter(!is.na(Q21), Q21 > 0) |>
  group_by(voter_category, Q21) |>
  summarize(weighted_count = sum(weight), .groups = "drop_last") |>
  mutate(
    weighted_total = sum(weighted_count),
    percentage = weighted_count / weighted_total * 100
  ) |>
  ungroup()

weighted_summary <- weighted_by_voter |>
  pivot_wider(
    id_cols = voter_category,
    names_from = Q21,
    values_from = percentage,
    names_prefix = "Q21_"
  ) |>
  rename(
    "Yes (%)" = Q21_1,
    "No (%)" = Q21_2,
    "Unsure (%)" = Q21_3
  )

kable(weighted_summary, digits = 1,
      caption = "WEIGHTED: Voting intentions by past voting behavior")
```

### Where Weighting Made the Biggest Difference

```{r category-differences}
differences <- unweighted_summary |>
  inner_join(weighted_summary, by = "voter_category", suffix = c("_unwtd", "_wtd")) |>
  mutate(
    yes_diff = `Yes (%)_wtd` - `Yes (%)_unwtd`,
    no_diff = `No (%)_wtd` - `No (%)_unwtd`,
    unsure_diff = `Unsure (%)_wtd` - `Unsure (%)_unwtd`
  ) |>
  select(voter_category, yes_diff, no_diff, unsure_diff)

kable(differences, digits = 1,
      col.names = c("Voter Category", "Yes (Δ)", "No (Δ)", "Unsure (Δ)"),
      caption = "Percentage point differences: Weighted minus Unweighted")
```

**Interpreting these numbers:**
- **Positive**: Weighted percentage is higher (group was under-represented in raw sample)
- **Negative**: Weighted percentage is lower (group was over-represented in raw sample)
- **Large absolute values**: Sample was very unrepresentative for this group

## Visualizing the Impact

Let's see this graphically:

```{r viz-comparison, fig.width=12, fig.height=6}
# Combine weighted and unweighted for visualization
combined <- bind_rows(
  unweighted_summary |> mutate(method = "Unweighted"),
  weighted_summary |> mutate(method = "Weighted")
) |>
  pivot_longer(
    cols = c("Yes (%)", "No (%)", "Unsure (%)"),
    names_to = "response",
    values_to = "percentage"
  )

ggplot(combined, aes(x = voter_category, y = percentage, fill = response)) +
  geom_col(position = "dodge", alpha = 0.8) +
  facet_wrap(~method) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Impact of Survey Weighting on Voting Intentions",
    subtitle = "Notice the shifts—especially among infrequent voters",
    x = "Voter Category",
    y = "Percentage",
    fill = "Response"
  ) +
  scale_fill_brewer(palette = "Set2")
```

The shifts might look subtle, but they're real and consequential.

## When Weighting Matters Most

Weighting has the biggest effect when:

**1. The sample is badly skewed**: You have too many of one demographic, too few of another

**2. Groups differ in their views**: The over-represented group has different opinions than the under-represented group

**3. Margins are close**: A 2-point shift can flip a "lead" into a "tie"

**4. Demographics correlate with outcomes**: Age predicts voting behavior, education predicts policy views, etc.

Weighting matters less when:
- Sample already matches population well
- All demographic groups have similar views
- Margins are very wide (60-30, not 48-46)

## Evaluating Polls as a Journalist

When a poll lands on your desk, here's your checklist:

**Ask these questions:**

✓ **Was the data weighted?** If not, be skeptical. Unweighted polls are almost never representative.

✓ **What variables were used for weighting?** Age, race, education, gender, past voting? More is usually better.

✓ **How large are the weights?** If max weight > 3.0, the sample was very unrepresentative. That's a red flag.

✓ **Is methodology disclosed?** Reputable pollsters explain their weighting in detail. Secrecy is a bad sign.

✓ **What was the response rate?** If only 5% of people respond, even perfect weighting can't fix a fundamentally biased sample.

**Red flags:**

🚩 No mention of weighting at all

🚩 Extreme weights (>5.0) suggest the sample is terrible

🚩 Weighting on the outcome itself (circular reasoning)

🚩 Vague methodology ("results were adjusted to reflect population")

🚩 Online opt-in polls with no probability sample

## How to Report Weighted Results

You've decided the poll is credible. How do you write about it?

**Bad version:**
"A new poll shows Johnson leading 54-43."

**Better version:**
"A new poll of 1,247 likely voters shows Mayor Johnson leading challenger Smith 54% to 43%, according to results weighted to match county demographics.

The poll, conducted by Reliable Polling Inc., contacted voters via a mix of landlines and cell phones. Results were weighted to reflect the county's age, gender, education, and racial composition, as well as past voting patterns.

Before weighting, Johnson's lead was 5 points wider, but adjustments to account for an overrepresentation of voters over age 65 narrowed the margin. The poll has a margin of error of ±3.2 percentage points."

**What this includes:**
- Sample size (1,247)
- How they were reached (landlines + cell phones)
- What variables were weighted (age, gender, education, race, past voting)
- That weighting changed results (5-point shift)
- Margin of error

## When Weighting Goes Wrong

Weighting can fix some problems, but it has limits:

**Problem 1: Over-reliance on small groups**

If you have only 10 respondents from a group that should be 15% of the sample, each one gets a weight around 1.5. But those 10 people might not represent their group well. You're amplifying a potentially unrepresentative subsample.

**Problem 2: Hiding a fundamentally bad sample**

Extreme weights (>3.0) mean your sample is so skewed that weighting is duct tape on a broken machine. You can't fix a landline-only survey by giving young people 5x weights—you simply didn't reach young people.

**Problem 3: Model dependency**

Different weighting schemes produce different results. Weight on age and gender? One answer. Add education and race? Different answer. This is why methodology matters—it determines the outcome.

## The Bottom Line

Survey weighting is a necessary tool, not a suspicious trick. Every reputable poll uses it because samples are never perfect.

**As a journalist, you need to:**
- Understand why weighting exists (samples are always biased)
- Know how to interpret weighted results (sum weights, don't count people)
- Evaluate whether weighting was done well (check methodology)
- Report it transparently (explain what was weighted and how)
- Recognize its limits (can't fix a terrible sample)

**Key principles:**
- Weights < 1: Group is over-represented in sample
- Weights > 1: Group is under-represented in sample
- Large weights (>3): Sample quality is questionable
- Always use weighted results when analyzing survey data
- Always report that results are weighted

## Your Turn

Test your understanding:

1. A survey over-samples college graduates (40% of sample, but only 30% of population). Should their weights be above or below 1.0? By approximately how much?

2. You see a poll where one respondent has a weight of 8.5. What does this tell you about the sample quality?

3. Find a recent poll in the news. Does it report weighting methodology? What variables does it weight on? Are there any red flags?

## Looking Ahead

We've covered how to make a single sample representative through weighting. But what if you want to understand how multiple factors simultaneously affect an outcome? That's where **multiple regression** comes in—our final chapter. It's regression (which we learned in Chapter 7) but with many predictors at once, letting you control for multiple confounding variables simultaneously.

For now, remember: weighted surveys are good. Unweighted surveys are misleading. And surveys with terrible samples can't be saved by any amount of weighting.
